---
title: "52414 - lab 1"
author: "52414"
date: "10/5/2023"
output: html_document
---


# *Lab 1: Basic Data Wrangling and Plotting, Distributions*  
<br/><br/>  
  
### Submission Instructions  
  
- This lab will be submitted in pairs (if you don't have a pair, please contact us) via the submission link in moodle.


- Your final submission should include two files: an `Rmd` file (with your answers filled-in) and an `html` file that was generated automatically by knitting the `Rmd` file using knitr.  Name your files as `<ID1>_<ID2>.Rmd` and `<ID1>_<ID2>.html` (insert your ID numbers instead). 


- **Grading:** There are $8$ questions with overall $15$ sub-questions. Each sub-question is worth $6\frac{2}{3}$ points to the overall lab grade. 
The questions vary in length and difficulty level. It is recommended to start with the simpler and shorter questions. 
Points may be reduced for incorrect naming of files, missing parts and problems in knitting the `Rmd` file and general appearance of the report. 

- **Libraries:** The only allowed libraries are listed below  (**do not add additional libraries without permission from the course staff**):
```{r, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
library(tidyverse) # This includes dplyr, stringr, ggplot2, .. 
library(data.table)
library(rworldmap) # world map
library(ggthemes)
library(reshape2) # melt: change data-frame format long/wide
library(e1071) # skewness and kurtosis
library(rvest)
library(corrplot)
library(moments)
library(spatstat.geom)
```  
<br/><br/>

## Analysis of the World Democracy Index Dataset         
    
The [wikipedia/Democracy_Index](https://en.wikipedia.org/wiki/Democracy_Index) website hosts world-wide data on different measurements of democracy index for world countries. 
For more information about it, please visit [here](https://web.archive.org/web/20170320185156/http://pages.eiu.com/rs/eiu2/images/Democracy-Index-2012.pdf).    <br>
  
We will focus on analyzing the changes in the index in different countries, as well as the individual components comprising the index, and comparison to other datasets. 


### General Guidance
- Your solution should be submitted as a full report integrating text, code, figures and tables. 
For each question, describe first in the text of your solution what you're trying to do, then include the relevant code, 
then the results (e.g. figures/tables) and then a textual description of them. 

- In most questions the extraction/manipulation of relevant parts of the data-frame can be performed using commands from the `tidyverse` and `dplyr` R packages, such as `head`, `arrange`, `aggregate`, `group-by`, `filter`, `select`, `summaries`, `mutate` etc.

- When displaying tables, show the relevant columns and rows with meaningful names, and describe the results. 

- When displaying figures, make sure that the figure is clear to the reader, axis ranges are appropriate, labels for the axis, title and different curves/bars are displayed clearly (font sizes are large enough), a legend is shown when needed etc. 
Explain and describe in text what is shown in the figure. 

- It could be that in some cases data are missing (e.g. `NA`). Make sure that all your calculations (e.g. taking the maximum, average, correlation etc.) take this into account. Specifically, the calculations should ignore the missing values to allow us to compute the desired results for the rest of the values (for example, using the option `na.rm = TRUE` or `us = "complete.obs"`). 



### Questions: 

1. **Loading data and basic processing:**
    a. Load the  [democracy-index](https://en.wikipedia.org/wiki/Democracy_Index) html into an `R` object, 
    using the `rvest` package. <br>
    Next, extract the three tables shown in the web-page as `List by region`, `List by country` and `components` into three separate `R` data-frames. 
Display the top five rows of each table to check that they were loaded correctly. <br>
    b. Display in a new table the `top` five *countries* in terms of the democracy index in *2022*. Show only the country name and the democracy index. <br>
Repeat the same with the five `bottom` countries in *2022*. <br>
Repeat the same with the five top and bottom countries according to the *average* index value of all the $15$ years between *2006* and *2022* for which data is available in the table `List by country`.


2. **Plotting distributions of groups of countries:**
    a. Make one figure showing seven boxplots representing the distributions of the `democracy index` in *2022* of the different world regions given in the `List by country` table (each boxplot should represent the distribution of all countries within a specific region). <br>
    Next, for each region that has at least one outlier country, find and list all the outliers that appear in the boxplot. <br>
    (**Hint:** You may use the `boxplot.stats` command). <br>
    b. Make a figure showing density plots for the same distributions of the `democracy index` in *2022* in the seven different regions. Do the densities resemble to the Normal distribution? 
Compute the *mean*, *variance*, *skewness* and *kurtosis* for all the distributions, display them in a table and explain what they mean about the empirical distribution of the data. 


3. **Comparing countries and showing trends in democracy index:**
    a. Write a function that receives as input a data-frame, and a vector of country names (as strings). The function plots the values of the `democracy index` of these countries in different colors as a function of the year (from *2006* to *2022*), shown on the same graph as curves with different colors or symbols. Use meaningful axis and plot labels, and add an informative legend. Use the function and plot of the `democracy index` for five countries of your choice. <br>
    Use the same function for the table `List by region` where the seven region names as inserted as input instead of `countries`, to show changes in the world regions `democracy index` over time. <br>
    
    b. Divide the countries into eight separate groups (clusters) as follows: 
- First, countries whose index increased (one cluster) or decreased (another cluster) by at least $1.5$ points between *2006* and *2022*. 
- Second, countries whose index increased (one cluster) or decreased (another cluster) by between $0.75$ to $1.5$ points between *2006* and *2022*. 
- Next, countries that dropped by at least $0.75$ points after $2006$, and then recovered by at least $0.75$ points in $2022$ compared to the lowest drop.
- Similarly, countries that increased by at least $0.75$ points after $2006$, and then dropped by at least $0.75$ points in $2022$ compared to the highest point.
- Next, countries that had barely changed from $2006$ to $2022$, i.e. that the difference between their highest and lowest index was less than $0.5$ points. 
- Finally, all other countries. <br>
For each of the eight groups of countries, plot their changes using the function from 2.a. Describe the patterns you see in the different groups. 

**Remark:** Don't worry if some of the groups you get are large with countries with very similar colors, and/or a small graph panel due to a large legend.


4. **Change in category:** <br>
    For each of the four different regime types (`Full democracy`, `Flawed democracy`, `Hybrid regime`,  `Authoritarian`), use the countries democracy index data frame to estimate the probability of a country to go from one such a regime in $2006$ to each of the other four regimes in $2022$.
    Show the results (sixteen estimated probabilities) in a $4$-by-$4$ table, and also in a heatmap. <br>
    **Remarks:** Your estimates should simply be the empirical frequencies - for example, if $2$ out $20$ countries moved from  `Authoritarian` in $2006$ to `Hybrid regime` in $2022$, then get an estimate of $0.2$ for the probability of such a regime change). <br>
    Use the table `By regime type` from the democracy index webpage to determine the regime type  category based on the democracy index value. 


5. **Joining data from additional tables:**
    a. Load tables for the [gdp](https://en.wikipedia.org/wiki/List_of_countries_by_GDP_(PPP)_per_capita), [population size](https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population), [incarnation rates](https://en.wikipedia.org/wiki/List_of_countries_by_incarceration_rate) and [area](https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_area) web pages using the `rvest` library. <br>
    For each of the four web-page, extract the table for countries into an `R` data-frame. <br>
    Join the table of the democracy index by country with these four table, using the countries names for joining. 
    Display the top five rows of the joined table <br>
    b. Fit a simple linear regression model with the `democracy index` at $2022$ as the predictor and `GDP (PPP) per capita`  (use the CIA estimates) as the response, 
    and report the regression results. <br>
    Next, Make a scatter plot of the  the `GDP (PPP) per capita` (y-axis) vs. the `democracy index` at $2022$, with the fittedthe regression line. 
    Describe your results. <br>
    Repeat the same steps with `incarnation rate (per 100,000)` as the response 


6. **Empirical Cumulative Distribution Function (CDF):**
    a. Let $X$ be a random variable representing the `GDP (PPP) per capita` of a randomly selected *country* in 2022, 
    where countries are selected uniformly at random from all world countries. Compute and plot the empirical 
    CDF of $X$. <br>
    b. Let $Y$ be a random variable representing the `GDP (PPP) per capita` of a randomly selected *person* in the world in 2022, 
    where a person is selected uniformly at random from all world population. Compute and plot the empirical CDF for $Y$ and explain the differences from the distribution for $X$.
    **Remark:** Use the population size data to compute the empirical CDF for this case. It is possible to use the library `spatstat.geom`.
    c. Let $Z$ be a random variable representing the `GDP (PPP) per capita` of a randomly selected *person* in the world in 2022, 
    where the *location* of the person is selected uniformly at random from all the land area on earth. Compute and plot the empirical CDF for $Y$ and explain the differences from the distribution for $X$.
    Compare the median, and the $25\%$ and $75\%$ percentiles of $X,Y$ and $Z$. Are they similar or different? explain. 
    **Remark:** Use the countries land area (in $km^2$ or $mi^2$) to compute the empirical CDF for this case. You will need to parse the corresponding column to get the numerical data. 


7. **Displaying data on the world map:** <br>
    Use the `rworldmap` package to display the world map and color each country based on the `average democracy index` across the years from $2006$ to $2022$. 
    Describe the resulting map in a couple of sentences. <br>
    Next, repeat all parts above , but this time display in the map the `difference` in the index between $2022$ and $2006$. 

    **Guidance:** Use the `joinCountryData2Map` and `mapCountryData` commands to make the plots. 
    Keep countries with missing data in white. 


8. **Coponents of the Demography Index:** 
    a. Join the `components` table with the main table from the previous questions. Display the top five rows. 
    Next, compute the correlation between all pairs of the five democracy components (`Electoral process and pluralism`, `Functioning of government`,  `Political participation`, `Political culture` and `Civil liberties`), and plot the resulting $5$-by-$5$ correlations matrix in a heatmap. (It is possible to use the `corrplot` library). <br>
    b. Run multiple linear regression where the covariates are the different democracy sub-indices form the `components` table, and the response variable that you try to predict the `GDP (PPP) per capita` of each country. <br>
    Show a summary of your regression analysis. What coefficients are significant at significance level $\alpha=0.01$? <br>
    What countries are outliers? Display the five countries with the highest and lowest residuals in a table.  Can you think of other factors contributing to their
    high/low `GDP (PPP) per capita`?




**Good luck!**

**Solution:**  (Fill code, text, plots etc.)

1.a. Loading the data via URL connection:
```{r, cache=TRUE} 
democracy <- read_html("https://en.wikipedia.org/wiki/Democracy_Index")
all.tables = html_nodes(democracy, "table")  

# Use html_table to extract the individual tables from the all.tables object:
categories <- as.data.frame(html_table(all.tables[3], fill = TRUE))  # Example 

#based on the example we called all the relevant tables in order and names them accordingly 
list_by_region <- as.data.frame(html_table(all.tables[4], fill = TRUE))
list_by_region22 <- as.data.frame(html_table(all.tables[5], fill = TRUE))
list_by_country <- as.data.frame(html_table(all.tables[6], fill = TRUE))
components <- as.data.frame(html_table(all.tables[7], fill = TRUE))

# we'll call the top five of the required variances
head(categories)
head(list_by_region)
head(list_by_country)
head(components)
```

1.b. 
```{r, cache=TRUE} 
top <- select(list_by_country,Country,X2022) %>% arrange(desc(X2022)) %>% head() 
bottom <- select(list_by_country,Country,X2022) %>% arrange(X2022) %>% head() 


averages <- list_by_country %>%
  mutate(avg = rowMeans(select(., 5:19), na.rm = TRUE)) %>%
  select(Country, avg)

top_averages<- averages %>% arrange(desc(avg)) %>% head() 
top_averages

bottom_averages<- averages %>% arrange(avg) %>% head() 
bottom_averages
```

*The top five countries in terms of the democracy index in 2022 are: Norway, New Zeland, Iceland, Sweden, Finland and Danmark
*The bottom five countries in terms of the democracy index in 2022 are: Afghanistan, Myanmar, North Korea, Central African Republic, Syria and Democratic Republic of the Congo
*The top five countries according to the average index value of all the 15 years are: Norway, Iceland, Sweden, Danmark New Zeland, Finland. We can see that they are the same contries than the top five in 2022, just in a different order.
*The top five countries according to the average index value of all the 15 years are: North Korea, Chad, Central African Republic, Syria, Turkmenistan, Democratic Republic of the Congo.



2.a. 
```{r, cache=TRUE} 

ggplot(list_by_country, aes(x = Region, y = `X2022`, fill = Region)) +
  geom_boxplot() +
  xlab("Region") +
  ylab("Democracy Index") +
  ggtitle("Boxplots by Region") +
  theme(axis.title.x = element_text(margin = margin(t = 10)),
        axis.text.x = element_text(angle = 45, hjust = 1.1, vjust = 1),
        plot.title = element_text(face = "bold", hjust = 0.5))


outliers <-list_by_country %>% group_by(Region) %>% summarise(outliers = paste(boxplot.stats(`X2022`)$out, collapse = ", "))

outliers
```

*We can see that for the region of Asia and Australasia, this is the region that have the biggest difference between the minimum index and the maximum, a large step may indicate that the data within that group exhibit a wide range of values. It has a median of approximately 6.25, the lower percentile at approximately 3.75 and the upper percentile at approximately 7.25. The majority of countries in this region are Flawed Democracys. It does not have any outliers.
*For the Central and Eastern Europe region, the width of the box looks like the precedent region but their gap between the maximum and the minimum is smaller. It's median is approximately 6.25, the lower percentile at approximately 3.60 and the upper percentile at approximately 7.The majority of countries in this region are Flawed Democracy. It does not have any outliers. This region and the precedent are very alike.
*For the Latin America and the Caribbean, the width of the box is a lot smaller than the two precedents regions, the small width indicates that the values are clustered closely together and have limited variability, but the gap between the minimum index and the maximum is big.  It has a median of approximately 6.25, the lower percentile at approximately 5 and the upper percentile at approximately 7.1. The majority of countries in this region are Flawed Democracy. It does not have any outliers.
*For the Middle east and North Africa, the width of the box is smaller than the precedent region, the small width indicates that the values are clustered closely together and have limited variability, the gap between the minimum index and the maximum is also not big. It's median is the lowest of every regions and is approximately 3.1, the lower percentile at approximately 2.5 and the upper percentile at approximately 3.75. The majority of countries in this region are Authoritarian regimes. It does have one outlier at approximately 7.93.
*For the North America, the width of the box is the smaller than every other regions, that is because it contains only two countries, and these two countries look alike as a term of the type of regime. The median is approximately 8.2 and it lower percentile, upper percentile, maximum and minimum are all between 7.75 and 9.The two countries seem to be or Flawed democracy or Full democracy. It does not have any outliers.
*For the Sub-Saharan Africa, the width of the box looks like the firsts two regions, this suggests that there is considerable variability in the variable being measured within that region.  It's median is approximately 3.75, the lower percentile at approximately 3.1 and the upper percentile at approximately 5.3. The democracy indexes can go from 1.25 to 8. The majority of countries in this region are Authoritarian regimes. It does not have any outlier.
*For the Western Europe,  the width of the box is smaller than the precedent region, the small width indicates that the values are clustered closely together and have limited variability, the gap between the minimum index and the maximum is also not big, can go from 7.4 to 9.9. It's median is approximately 8, the lower percentile at approximately 7.8 and the upper percentile at approximately 8.9. The majority of countries in this region are Full democracy. This region looks a lot like the North America's region. It does have one outlier at approximately 4.35.

2.b. 
```{r, cache=TRUE} 
# Create density plot
ggplot(list_by_country, aes(x = `X2022`, fill = Region)) +
  geom_density(alpha = 0.2) +
  xlab("Democracy Index 2022") +
  ylab("Density") +
  ggtitle("Density Plot by Region") +
  theme(legend.position = "top")

# Compute summary statistics
summary_table <- list_by_country %>%
  group_by(Region) %>%
  summarize(
    Mean = mean(`X2022`),
    Variance = var(`X2022`),
    Skewness = moments::skewness(`X2022`),
    Kurtosis = moments::kurtosis(`X2022`)
  ) 

summary_table
```


*Skewness is a statistical number that tells us if a distribution is symmetric or not.
We can see on the table that we have created that the skewness or the distribution of north America is 0,therefor its distribution is symmetric.
If a distribution is symmetric its a normal distribution, therefor its median, mean and mode are equals.
f Skewness is greater than 0, then it is called right-skewed or that the right tail is longer than the left tail. If Skewness is less than 0, then it is called left-skewed or that the left tail is longer than the right tail.

The right-skewed regions are Middle East and North Africa and Sub-Saharan Africa.
The left-skewed regions are Asia and Australasia,Central and Eastern Europe, Latin America and the Caribbean and Western Europe.

*Kurtosis is a statistical number that tells us if a distribution is taller or shorter than a normal distribution. If a distribution is similar to the normal distribution, the Kurtosis value is 0. If Kurtosis is greater than 0, then it has a higher peak compared to the normal distribution. If Kurtosis is less than 0, then it is flatter than a normal distribution.

In our case all the regions Kurtosis are greater than 0, therefor they all have a higher peak compared to the normal distribution.

As we saw in the precedent question, the region Asia and Australasia has a big variance and on the contrary the region North America has a very low variance. Furthermore, we can conclude that in average the countries of the regions Asia and Australasia, Central and Eastern Europe, Latin America and the Caribbean and Sub-Saharan Africa have an hybrid regime; on average the countries in the Middle East and North Africa have an Authoritarian regime and the countries in North America and Western Europe have on average Full democracies.


3.a. 
```{r, cache=TRUE} 
Comparing_countries <- function(data, country_names) {
  
  if ("Country" %in% colnames(data)) {
    place_col <- "Country"
  } else {
    place_col <- "Region"
  } 
   
  year_cols <- grep("^X[0-9]{4}$", colnames(data), value = TRUE)
  years <- unique(as.numeric(substr(year_cols, 2,5)))

  fil_data <- data[data[[place_col]] %in% country_names, ]
  color <- rainbow(length(country_names))

  plot(NULL, xlim = range(years), ylim = c(0, 10),
       xlab = "Year", ylab = "Democracy Index",
       main = "Democracy Index for Countries")  
 
  axis(1, at = years, labels = years)

  for (i in 1:length(country_names)) {
    place <- country_names[i]
    country_data <- fil_data[fil_data[[place_col]] == place, ]
    lines(years, t(country_data[year_cols]), type = "l", col = color[i])
  }
  
  legend("bottomright", inset = 0.02, legend = country_names,
         col = color, pch = 19, bty = "n")

}
   

Comparing_countries(list_by_region, c("Asia and Australasia", "Sub-Saharan Africa"))

```

YOUR ANALYSIS HERE


3.b. 
```{r, cache=TRUE} 
index_diff <- list_by_country$X2022 - list_by_country$X2006
#first cluster
big_increase_cluster <- list_by_country[index_diff >= 1.5, "Country"]
#second cluster
big_decrease_cluster <- list_by_country[index_diff <= -1.5, "Country"]
#third cluster
small_increase_cluster <- list_by_country[index_diff > 0.75 & index_diff <= 1.5, "Country"]
#fourth cluster
small_decrease_cluster <- list_by_country[index_diff < -0.75 & index_diff >= -1.5, "Country"]

list_by_country$min_index <- apply(list_by_country[, 6:19], 1, min)

#fifth cluster
index_min_diff_2006 <- list_by_country$X2006 - list_by_country$min_index
index_min_diff_2022 <- list_by_country$X2022 - list_by_country$min_index

decrease_increase <- list_by_country$Country[index_min_diff_2006 >= 0.75 & index_min_diff_2022 >= 0.75]

#sixth cluster
list_by_country$max_index <- apply(list_by_country[, 6:19], 1, max)
index_max_diff_2006 <- list_by_country$X2006 - list_by_country$max_index
index_max_diff_2022 <- list_by_country$X2022 - list_by_country$max_index
increase_decrease <- list_by_country$Country[index_max_diff_2006 <= -0.75 & index_max_diff_2022 <= -0.75]

#seventh cluster
bare_change <- list_by_country$Country[list_by_country$max_index-list_by_country$min_index < 0.5 ]

#height cluster
other_countries <- list_by_country$Country[!(list_by_country$Country %in% c(big_increase_cluster,big_decrease_cluster,small_increase_cluster, small_decrease_cluster, decrease_increase, increase_decrease, bare_change))]

#comparing countries

Comparing_countries(list_by_country, big_increase_cluster)
Comparing_countries(list_by_country, big_decrease_cluster)
Comparing_countries(list_by_country, small_increase_cluster)
Comparing_countries(list_by_country, small_decrease_cluster)
Comparing_countries(list_by_country, decrease_increase)
Comparing_countries(list_by_country, increase_decrease)
Comparing_countries(list_by_country, bare_change)
Comparing_countries(list_by_country, other_countries)
```

YOUR ANALYSIS HERE

4.
```{r, cache=TRUE} 
# Step 1: Filter the data table for 2006 and 2022
data_for_freq <- select(list_by_country,Country,X2006,X2022)


data_for_freq$Regime_Type_2006 <- NA
data_for_freq$Regime_Type_2022 <- NA

# Assign regime type based on score conditions
data_for_freq$Regime_Type_2006 <- ifelse(data_for_freq$X2006 >= 8.01 & data_for_freq$X2006 <= 10, "Full democracies",
                              ifelse(data_for_freq$X2006 >= 6.01 & data_for_freq$X2006 <= 8, "Flawed democracies",
                              ifelse(data_for_freq$X2006 >= 4.01 & data_for_freq$X2006 <= 6, "Hybrid regimes",
                              ifelse(data_for_freq$X2006 >= 0 & data_for_freq$X2006 <= 4, "Authoritarian regimes", NA))))

data_for_freq$Regime_Type_2022 <- ifelse(data_for_freq$X2022 >= 8.01 & data_for_freq$X2022 <= 10, "Full democracies",
                              ifelse(data_for_freq$X2022 >= 6.01 & data_for_freq$X2022 <= 8, "Flawed democracies",
                              ifelse(data_for_freq$X2022 >= 4.01 & data_for_freq$X2022 <= 6, "Hybrid regimes",
                              ifelse(data_for_freq$X2022 >= 0 & data_for_freq$X2022 <= 4, "Authoritarian regimes", NA))))

data_for_freqs <- table(data_for_freq$Regime_Type_2006 ,data_for_freq$Regime_Type_2022)

transition_probabilities <- prop.table(data_for_freqs, margin = 1)
transition_probabilities

regime_categories <- c("Full democracy", "Flawed democracy", "Hybrid regime", "Authoritarian")
prob_table <- matrix(transition_probabilities, nrow = 4, byrow = TRUE, dimnames = list(regime_categories, regime_categories))

print(prob_table)

# Create a heatmap
heatmap(prob_table, col = colorRampPalette(c("white", "green"))(20), main = "Regime Transition Probabilities")
```

YOUR ANALYSIS HERE



5.a. 
```{r, cache=TRUE}
gdp_url<- read_html("https://en.wikipedia.org/wiki/List_of_countries_by_GDP_(PPP)_per_capita")
population_size_url<- read_html("https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_population")
incarnation_rates_url<- read_html("https://en.wikipedia.org/wiki/List_of_countries_by_incarceration_rate")
area_url<- read_html("https://en.wikipedia.org/wiki/List_of_countries_and_dependencies_by_area")

gdp.tables = html_nodes(gdp_url, "table")  
gdp_table <- as.data.frame(html_table(gdp.tables[2], fill = TRUE))  


pop.tables = html_nodes(population_size_url, "table")  
population_size_table <- as.data.frame(html_table(pop.tables[2], fill = TRUE))  

incar.tables = html_nodes(incarnation_rates_url, "table")  
incarnation_rate_table <- as.data.frame(html_table(incar.tables[2], fill = TRUE))


area.tables = html_nodes(area_url, "table")  
area_table <- as.data.frame(html_table(area.tables[2], fill = TRUE))  

colnames(gdp_table)[colnames(gdp_table) == "Country.Territory"] <- "Country"
colnames(population_size_table)[colnames(population_size_table) == "Country...Dependency"] <- "Country"
colnames(incarnation_rate_table)[colnames(incarnation_rate_table) == "Location"] <- "Country"
colnames(area_table)[colnames(area_table) == "Country...Dependency"] <- "Country"
gdp_table$Country <- gsub("\\*$", "", gdp_table$Country)
gdp_table$Country <- gsub("\\ ", "", gdp_table$Country)


joined_table <- merge(list_by_country, gdp_table, by = "Country", all.x = TRUE)
joined_table <- merge(joined_table, population_size_table, by = "Country", all.x = TRUE)
joined_table <- merge(joined_table, incarnation_rate_table, by = "Country", all.x = TRUE)
joined_table <- merge(joined_table, area_table, by = "Country", all.x = TRUE)
joined_table <- as.data.frame(joined_table)

head(joined_table)
```

YOUR ANALYSIS HERE


5.b. 
```{r, cache=TRUE} 
joined_table <- data.frame(lapply(joined_table, function(x) gsub(",", "", x)))
colnames(joined_table)[colnames(joined_table) == "IMF.5..6."] <- "IMF.est"
colnames(joined_table)[colnames(joined_table) == "IMF.5..6..1"] <- "IMF.year"
colnames(joined_table)[colnames(joined_table) == "World.Bank.7."] <- "World.bank.est"
colnames(joined_table)[colnames(joined_table) == "World.Bank.7..1"] <- "World.bank.year"
colnames(joined_table)[colnames(joined_table) == "CIA.8..9..10."] <- "CIA.est"
colnames(joined_table)[colnames(joined_table) == "CIA.8..9..10..1"] <- "CIA.year"

data <- na.omit(joined_table[, c("CIA.est", "X2022")])

data$CIA.est <- as.numeric(data$CIA.est)
data$X2022 <- as.numeric(data$X2022)

reg_mod_cia <- lm(CIA.est ~ X2022, data = data)
summary(reg_mod_cia)

plot(data$X2022, data$CIA.est, xlab = "Democracy Index", ylab = "GDP (PPP) per capita")
abline(reg_mod_cia, col = "red")

colnames(joined_table)[colnames(joined_table) == "Rate.per.100.000..3."] <- "Rate.100000"
data1 <- na.omit(joined_table[, c("Rate.100000", "X2022")])

data1$Rate.100000 <- as.numeric(data1$Rate.100000)
data1$X2022 <- as.numeric(data1$X2022)

reg_mod_inca <- lm(Rate.100000 ~ X2022, data = data1)
summary(reg_mod_inca)

plot(data1$X2022, data1$Rate.100000, xlab = "Democracy Index", ylab = "Incarnation Rate")
abline(reg_mod_inca, col="red")


```

YOUR ANALYSIS HERE

6.a. 
```{r, cache=TRUE} 
sampled_gdp <- sample(joined_table$X2022, size = 1000, replace = TRUE)

# Compute empirical CDF
empirical_cdf <- ecdf(sampled_gdp)

# Generate a sequence of values
x <- seq(min(sampled_gdp), max(sampled_gdp), length.out = 100)

# Calculate empirical CDF values for the sequence of x values
y <- empirical_cdf(x)

# Create a data frame with x and y values
cdf_data <- data.frame(x, y)

# Plot the empirical CDF
ggplot(cdf_data, aes(x = x, y = y)) +
  geom_step() +
  labs(x = "GDP (PPP) per capita", y = "Empirical CDF") +
  ggtitle("Empirical CDF of GDP (PPP) per capita for a Randomly Selected Country")
```

YOUR ANALYSIS HERE


6.b. 
```{r, cache=TRUE} 
library(ggplot2)

joined_table$Population <- as.numeric(joined_table$Population)

# Randomly sample GDP values for a selected person
sampled_gdp_person <- sample(joined_table$X2022, size = 55, replace = TRUE)

# Filter out non-numeric and missing values from GDP
filtered_gdp_person <- na.omit(as.numeric(sampled_gdp_person))

# Calculate the cumulative sum of the population size
population_cdf <- cumsum(joined_table$Population) / sum(joined_table$Population)

# Create a data frame with filtered GDP values and population CDF
cdf_data_person <- data.frame(GDP = filtered_gdp_person, PopulationCDF = population_cdf)

# Sort the data frame by GDP values
cdf_data_person <- cdf_data_person[order(cdf_data_person$GDP), ]

# Plot the empirical CDF for GDP and population
ggplot() +
  geom_step(data = cdf_data_person, aes(x = GDP, y = PopulationCDF), color = "red", linewidth = 1) +
  labs(x = "GDP (PPP) per capita", y = "Empirical CDF", title = "Empirical CDF of GDP (PPP) per capita") +
  scale_x_log10() +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal()
```

YOUR ANALYSIS HERE


6.c. 
```{r, cache=TRUE} 
colnames(joined_table)[colnames(joined_table) == "Landin.km2..mi2."] <- "LandArea"
land_area <- as.numeric(joined_table$LandArea)

# Remove non-numeric and missing values from X2022 column
x2022_numeric <- as.numeric(joined_table$X2022)
x2022_clean <- x2022_numeric[!is.na(x2022_numeric)]

# Sample GDP values for a person
sampled_gdp_person <- sample(x2022_clean, size = 55, replace = TRUE)

# Sample GDP values for land area
sampled_gdp_land_area <- sample(x2022_clean, size = 55, replace = TRUE)

# Calculate the cumulative distribution of land area
land_area_cdf <- cumsum(land_area) / sum(land_area)

# Calculate the empirical CDF for the sampled GDP values based on person
empirical_cdf_person <- ecdf(sampled_gdp_person)

# Generate a sequence of GDP values
gdp_values <- seq(min(sampled_gdp_person), max(sampled_gdp_person), length.out = 100)

# Calculate the empirical CDF for the sampled GDP values based on land area
empirical_cdf_land_area <- ecdf(sampled_gdp_land_area)

# Calculate the cumulative probabilities using the empirical CDF based on land area
cdf_values_land_area <- empirical_cdf_land_area(gdp_values)

# Create a data frame with GDP values and their corresponding cumulative probabilities based on land area
cdf_data_land_area <- data.frame(GDP = gdp_values, Probability = cdf_values_land_area)

# Plot the empirical CDF for Z
ggplot(cdf_data_land_area, aes(x = GDP, y = Probability)) +
  geom_step() +
  labs(x = "GDP (PPP) per capita", y = "Empirical CDF") +
  ggtitle("Empirical CDF of GDP (PPP) per capita for a Randomly Selected Person (Uniformly from Land Area)")

# Calculate the median, 25th percentile, and 75th percentile for X, Y, and Z
percentiles <- c(0.25, 0.5, 0.75)
x_percentiles <- quantile(x2022_clean, percentiles)
y_percentiles <- quantile(sampled_gdp_person, percentiles)
z_percentiles <- quantile(sampled_gdp_land_area, percentiles)

# Print the results
cat("X - Median:", x_percentiles[2], "\n")
cat("X - 25th Percentile:", x_percentiles[1], "\n")
cat("X - 75th Percentile:", x_percentiles[3], "\n\n")

cat("Y - Median:", y_percentiles[2], "\n")
cat("Y - 25th Percentile:", y_percentiles[1], "\n")
cat("Y - 75th Percentile:", y_percentiles[3], "\n\n")

cat("Z - Median:", z_percentiles[2], "\n")
cat("Z - 25th Percentile:", z_percentiles[1], "\n")
cat("Z - 75th Percentile:", z_percentiles[3], "\n")
```

YOUR ANALYSIS HERE


7 
```{r, cache=TRUE} 
#select the country names and from year 2006 till 2022
avg_table <- select(list_by_country,Country,c(5:19))

#create a new empty column to store the average for each country 
avg_table$avg <- NA
row_sums <- rowSums(avg_table[, c(2:16)])
#enter and calculate averages for each country 
avg_table$avg <- row_sums/15

world_map <- joinCountryData2Map(avg_table, joinCode = "NAME", nameJoinColumn = "Country")

#The world map with color based on average value
mapCountryData(world_map, nameColumnToPlot = "avg", mapTitle = "Average Value",
               catMethod = "fixedWidth", numCats = 10, missingCountryCol = "white", addLegend = TRUE, oceanCol = "lightblue")



table_2006 <- select(list_by_country,Country,c(19))
mapCountryData(world_map, nameColumnToPlot = "X2006…
[16:39, 30/05/2023] Lisa Stat : 5.bjoined_table <- data.frame(lapply(joined_table, function(x) gsub(",", "", x)))
colnames(joined_table)[colnames(joined_table) == "IMF.5..6."] <- "IMF.est"
colnames(joined_table)[colnames(joined_table) == "IMF.5..6..1"] <- "IMF.year"
colnames(joined_table)[colnames(joined_table) == "World.Bank.7."] <- "World.bank.est"
colnames(joined_table)[colnames(joined_table) == "World.Bank.7..1"] <- "World.bank.year"
colnames(joined_table)[colnames(joined_table) == "CIA.8..9..10."] <- "CIA.est"
colnames(joined_table)[colnames(joined_table) == "CIA.8..9..10..1"] <- "CIA.year"

reg_mod_cia <- lm(`CIA.est` ~ X2022, data = na.omit(joined_table[, c("CIA.est", "X2022")]))
summary(reg_mod_cia)

ggplot(na.omit(joined_table[, c("CIA.est", "X2022")]), aes(x = X2022, y = CIA.est)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(x = "Democracy Index 2022", y = "GDP (PPP) per capita") +
  ggtitle("Scatter Plot of GDP vs. Democracy Index with Regression Line")


colnames(joined_table)[colnames(joined_table) == "Rate.per.100.000..3."] <- "Rate.100000"
reg_mod_inca <- lm(`Rate.100000` ~ X2022, data = na.omit(joined_table[, c("Rate.100000", "X2022")]))
summary(reg_mod_inca)

ggplot(na.omit(joined_table[, c("Rate.100000", "X2022")]), aes(x = X2022, y = Rate.100000)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(x = "Democracy Index 2022", y = "Rate incarceration (per 100,000)") +
  ggtitle("Scatter Plot of Rate Incarceration vs. Democracy Index with Regression Line")
```

YOUR ANALYSIS HERE




8.a. 
```{r, cache=TRUE} 
#Joining the components table in a new variable:


joined_comp <- merge(components,joined_table,by = "Country")

#changing the titles of 5 columns that we'll need :

colnames(joined_comp)[colnames(joined_comp) == "Elec.toral.pro.cessand.plura.lism"] <- "Electoral_processand_pluralism"
colnames(joined_comp)[colnames(joined_comp) == "Func.tioningof.govern.ment"] <- "Functioning_of_government"
colnames(joined_comp)[colnames(joined_comp) == "Poli.ticalpartici.pation"] <- "Political_participation"
colnames(joined_comp)[colnames(joined_comp) == "Poli.ticalcul.ture"] <- "Political_culture"
colnames(joined_comp)[colnames(joined_comp) == "Civilliber.ties"] <- "Civil_liberties"

#Show the first 5
head(joined_comp,5)

#calculate the correlations between those 

selected_columns <- select(joined_comp,Electoral_processand_pluralism, Functioning_of_government, 
                      Political_participation, Political_culture, Civil_liberties)

selected_columns <- as.data.frame(lapply(selected_columns, as.numeric))

cor_matrix <- cor(selected_columns)

corrplot(cor_matrix, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45)
```

YOUR ANALYSIS HERE


8.b. 
```{r, cache=TRUE} 

# Fit the multiple linear regression model
reg_model <- lm(CIA.est ~ Electoral_processand_pluralism + Functioning_of_government + Political_participation + Political_culture + Civil_liberties, data = joined_comp)

# Display the summary of the regression analysis
summary(reg_model)
```

YOUR ANALYSIS HERE
